Vectorizing Python Constructs Using Pythran and Boost SIMD

### 2013-11-29 doc/papers/wpmvp14/paper.tex
Update experiments

>>>>

To validate the approach, a number of code snippets presenting vectorization
opportunities have been selected and benchmarked. The original Python version
is used as a reference, then code is compiled with Pythran without
vectorization support and finally with Pythran and vectorization support. Time
measurements are made using the \texttt{timeit} module.

Experiments are run on a computer running a Linux kernel version 3.10. The
processors are Intel Core i7, so the AVX instruction set is available. Python
version is 2.7.5, Pythran version is the \texttt{HEAD} of the \texttt{wpmvp14}
branch of the official Pythran repository
\url{https://github.com/serge-sans-paille/pythran}. The backend C++ compiler is
Clang version 3.4-1. The compilation flags used are \texttt{-O2} without
vectorisation support and \texttt{-O2 -march=native -DUSE\_BOOST\_SIMD} with
vectorisation support. Using \texttt{-O3} or \texttt{-Ofast} actually degraded
performances.

\subsection{Vectorization of the \texttt{map} intrisinc}

To evaluate the vectorization of the \texttt{map} intrisinc, two expressions
are used: \texttt{map(abs, r)} and \texttt{map(lambda x: 5 * x + 3, r)}, where
\texttt{r} is a \texttt{ndarray} of random double-precision floating point
numbers. Table~\ref{tbl:map-timings} summarizes the result. The execution time
for the Cython version is extremly long due to the ndarray indexing and
function call overheads in CPython. Thanks to static typing and aggressive
inlining, Pythran does not suffer from these issues, as shown by its sequential
execution time. Vectorization is beneficial and yields a $\times2.93$ speedup.

\begin{table}

    \centering
    \begin{tabular}{|l|r|r|r|}
        \hline
                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
        \hline
        \texttt{map(abs, r)}                    &   1530ms  & 1.72ms    & 0.587ms   \\
        \hline
        \texttt{map(f, r)}    &   141ms   & 1.72ms    & 0.587ms  \\
        \hline

    \end{tabular}
    \caption{Timings of various \texttt{map} calls using CPython, Pythran, and Pythran with vectorization.}
    \label{tbl:map-timings}

\end{table}

\subsection{Vectorization of the \texttt{sum} intrisinc}

To evaluate the vectorization of the \texttt{sum} intrisinc, two expressions
are used: \texttt{sum(r)} and \texttt{sum(r * r)}, where \texttt{r} is a
\texttt{ndarray} of random double-precision floating point numbers.
Table~\ref{tbl:sum-timings} summarizes the result. Again, CPython is extremely
slow. It is interesting to note that the extra array multiplication does not
impacts the excution time much, simply because it is an operation pefromed at
the native level by the \texttt{numpy} module, while the \texttt{sum} is fully
interpreted. The vectorized reductions yields a decent $\times2.41$ speedup
over the sequential version.

\begin{table}

    \centering
    \begin{tabular}{|l|r|r|r|}
        \hline
                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
        \hline
        \texttt{sum(r)}                    &   277ms  & 1.07ms    & 0.416ms   \\
        \hline
        \texttt{sum(r * r)}    &   292ms   & 1.09ms    & 0.452ms  \\
        \hline

    \end{tabular}
    \caption{Timings of various \texttt{sum} calls using CPython, Pythran, and Pythran with vectorization.}
    \label{tbl:sum-timings}

\end{table}

\subsection{Vectorization of \texttt{ndarray} point-to-point operations}

To evaluate the vectorization of \texttt{ndarray} point-to-point operations,
three expressions are used: \texttt{1. / a + a}, the example from
Listing~\ref{lst:numpy-code} and a computation of the point-to-point arc
distance  between random values. Again, all inputs are large randomized
\texttt{ndarray} of double precision floating point numbers.
Table~\ref{tbl:numpy-timings} summarizes the results. Because all computations
are done in native space, there is not much speedup when running them through
Pythran. The only benefits comes from the elimination of temporary arrays and
loop fusion. As a consequence, the more elements in the \texttt{numpy}
expression, the more Pythran accelerates the code. The vectorization speedups
follow the same process, as larger expression tend to increase the operation
over memory movement ratio.

\begin{table}

    \centering
    \begin{tabular}{|l|r|r|r|}
        \hline
                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
        \hline
        \texttt{1. / a + a}                    &   7.48ms  & 6.19ms    & 4.05ms   \\
        \hline
        \emph{rosen}                           &   13.9ms   & 2.98ms   & 1.15ms  \\
        \hline
        \emph{arc distance}                    &   178ms   & 146ms    & 53.8ms  \\
        \hline

    \end{tabular}
    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
    \label{tbl:numpy-timings}

\end{table}

\subsection{Vectorization of an Euler Problem}

This experiment considers the code from Listing~\ref{lst:euler06-py}.

\begin{table}

    \centering
    \begin{tabular}{|l|r|r|r|}
        \hline
                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
        \hline
        \emph{euler06}                    &   12ms   & 0.34ms    & 0.30ms  \\
        \hline

    \end{tabular}
    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
    \label{tbl:numpy-timings}

\end{table}
